### 2015-03-02 src/4_OnlineLearning.tex
efficient computation

%\subsection{Kernel methods}
%Kernel functions may be used to introduce a nonlinear separation
%in input space despite the linear classifier in decision space

\section{Online learning} \label{sec:onoffline}

@@

\begin{equation}
\label{eq:AARexpand}
\mathbf{X}_{aar} = \displaystyle \big (\sum_{t=1}^{m+1}
\mathbf{a}_t \mathbf{a}_t  ^\intercal + \gamma \mathbb{I}\big )^{-1}
\sum_{t=1}^m \mathbf{a}_t \mathbf{b}_t \, .
\end{equation}

If we define $\displaystyle\mathbf{S}= \sum_{t=1}^{m+1} \mathbf{a}_t
\mathbf{a}_t  ^\intercal + \gamma \mathbb{I} $ and $\mathbf{W}=
\displaystyle\sum_{t=1}^m \mathbf{a}_t \mathbf{b}_t$, the
algorithm~\ref{alg:AAR} is slightly different to the algorithm~\ref{alg:RR}, 
which updated matrix $\mathbf{S}$ before making the prediction:

\begin{algorithm}[ht]
\begin{algorithmic}[1]
\REQUIRE $\,$ \\
$\{\mathbf{a}_1,\dots,\mathbf{a}_{m} \}$: $m$ input vectors \\
$\{\mathbf{b}_1,\dots,\mathbf{b}_{m} \}$: $m$ target vectors \\
$\lambda$: regularization parameter \\
\ENSURE  $\,$ \\
$\{f(\mathbf{a}_1),\dots,f(\mathbf{a}_{m}) \}$: model predictions \\
\STATE Initialize $\mathbf{S}=\lambda \mathbb{I}$
and $\mathbf{W}=0$
\FOR { $t = 1$ to $m$ }
	\STATE read new $\mathbf{a}_t$
   	\STATE $\mathbf{S} = \mathbf{S} + \mathbf{a}_t \mathbf{a}_t^\intercal$
	\STATE $\mathbf{X}=\mathbf{S}^{-1}\mathbf{W}$
	\STATE output prediction $f(\mathbf{a}_t) = \mathbf{X}^\intercal \mathbf{a}_t$
   	\STATE Read new $\mathbf{y}_t$
    	\STATE $\mathbf{W} = \mathbf{W} + \mathbf{a}_t \mathbf{b}_t$
\ENDFOR
\end{algorithmic}
\caption{{\em The aggregating algorithm for regression}}
\label{alg:AAR}
\end{algorithm}

>>>>

%\subsection{Kernel methods}
%Kernel functions may be used to introduce a nonlinear separation
%in input space despite the linear classifier in decision space
\newpage
\section{Online learning} \label{sec:onoffline}

@@

\begin{equation}
\label{eq:AARexpand}
\mathbf{X}_{aar} = \displaystyle \big (\sum_{t=1}^{m+1}
\mathbf{a}_t \mathbf{a}_t  ^\intercal + \gamma \mathbb{I}\big )^{-1}
\sum_{t=1}^m \mathbf{a}_t \mathbf{y}_t \, .
\end{equation}

If we define $\displaystyle\mathbf{S}= \sum_{t=1}^{m+1} \mathbf{a}_t
\mathbf{a}_t  ^\intercal + \gamma \mathbb{I} $ and $\mathbf{W}=
\displaystyle\sum_{t=1}^m \mathbf{a}_t \mathbf{y}_t$, the
algorithm~\ref{alg:AAR} is slightly different to the algorithm~\ref{alg:RR}, 
which updated matrix $\mathbf{S}$ before making the prediction:

\begin{algorithm}[ht]
\begin{algorithmic}[1]
\REQUIRE $\,$ \\
$\{\mathbf{a}_1,\dots,\mathbf{a}_{m} \}$: $m$ input vectors \\
$\{\mathbf{y}_1,\dots,\mathbf{y}_{m} \}$: $m$ target vectors \\
$\lambda$: regularisation parameter \\
\ENSURE  $\,$ \\
$\{f(\mathbf{a}_1),\dots,f(\mathbf{a}_{m}) \}$: model predictions \\
\STATE Initialize $\mathbf{S}=\lambda \mathbb{I}$
and $\mathbf{W}=0$
\FOR { $t = 1$ to $m$ }
	\STATE read new $\mathbf{a}_t$
   	\STATE $\mathbf{S} = \mathbf{S} + \mathbf{a}_t \mathbf{a}_t^\intercal$
	\STATE $\mathbf{X}=\mathbf{S}^{-1}\mathbf{W}$
	\STATE output prediction $f(\mathbf{a}_t) = \mathbf{X}^\intercal \mathbf{a}_t$
   	\STATE Read new $\mathbf{y}_t$
    	\STATE $\mathbf{W} = \mathbf{W} + \mathbf{a}_t \mathbf{y}_t$
\ENDFOR
\end{algorithmic}
\caption{{\em The aggregating algorithm for regression}}
\label{alg:AAR}
\end{algorithm}

\section{Efficient computation}

In order to solve OLS in a efficient way, Coleman and
Sun~\cite{coleman+sun2010} presented an iterative algorithm which
uses $\mathbf{X}(\lambda)$ to approximate $\mathbf{X}(0)$.

Using the compact SVD (shown in equation~(\ref{eq:compactsvd}))
$\mathbf{X}(\lambda)$ is expressed as follows:

\begin{eqnarray}
\label{eq:optsolRRsvd}
\mathbf{X}(\lambda) & = & (\mathbf{A}^\top \mathbf{A}+ \lambda
\mathbb{I})^{-1}\mathbf{A}^\top \mathbf{Y} \nonumber \\
& = &\mathbf{V}_1(\Sigma_1^2+\lambda \mathbb{I})^{-1}\mathbf{\Sigma_1
U_1^\top Y}
\end{eqnarray}

\noindent where is easy to see that $\mathbf{X}(\lambda) \rightarrow
\mathbf{\hat{X}}=\mathbf{V}_1 \mathbf{\Sigma}_1^{-1}\mathbf{U}_1^\top
\mathbf{Y}$ as $\lambda \rightarrow 0$.

The method consists in obtaining $\mathbf{X}(\lambda)$ and then refine
by adding more terms of its Taylor expansion to approximate
$\mathbf{X}(0) = \mathbf{\hat{X}}$. The Taylor expansion about $\lambda_0$ is:

\begin{equation}
\label{eq:taylor}
    \mathbf{X}(\lambda)=\mathbf{X}(\lambda_0) + \sum_{k=1}^\infty
    \mathbf{s}_k(\lambda-\lambda_0)^{k}
\end{equation}

\noindent where $\mathbf{s}_k=\frac{1}{k!}\mathbf{X}(\lambda)^{(k)}$
and $\mathbf{X}(\lambda)^{(k)}$ is obtained by taking differences 
$\frac{\partial}{\partial \lambda}$ of:

\begin{eqnarray*}
(\mathbf{A}^\top \mathbf{A}+ \lambda\mathbb{I}) \mathbf{X}(\lambda) & = & \mathbf{A}^\top \mathbf{Y}\\
(\mathbf{A}^\top \mathbf{A}+ \lambda\mathbb{I}) \mathbf{X}(\lambda)^{(1)} + \mathbf{X}(\lambda)& = & 0 \\
\mathbf{X}(\lambda)^{(1)}  &=& -(\mathbf{A}^\top \mathbf{A}+ \lambda\mathbb{I}) ^{-1} \mathbf{X}(\lambda) \\
\mathbf{X}(\lambda)^{(2)}  &=& -2(\mathbf{A}^\top \mathbf{A}+ \lambda\mathbb{I}) ^{-1} \mathbf{X}(\lambda)^{(1)} \\
& \vdots & \\
\mathbf{X}(\lambda)^{(k)}  &=& -k(\mathbf{A}^\top \mathbf{A}+ \lambda\mathbb{I}) ^{-1} \mathbf{X}(\lambda)^{(k-1)} 
\end{eqnarray*}

\noindent for $\lambda=0$ we have:

\begin{equation}
\label{eq:taylor}
    \mathbf{X}(0)=\mathbf{X}(\lambda_0) + \sum_{k=1}^\infty
     (-1)^k \mathbf{s}_k \lambda_0^k
\end{equation}

\noindent therefore in order to ensure convergence, we can see that$\lambda_0$
selection cannot be large.

The algorithm for computing $\mathbf{X}(0)$ is the following:

\begin{algorithm}[H]
\begin{algorithmic}[1]
\REQUIRE $\,$ \\
$\mathbf{A}$: design matrix \\
$\mathbf{Y}$: response matrix \\
$\lambda$: rank deficient parameter \\
\ENSURE  $\,$ \\
$\mathbf{X}$: parameters \\
\STATE $\mathbf{M}=\mathbf{A^\top A}$ \\
\STATE Initialize $\mathbf{Q R}=\mathbf{M}$ \\
\STATE $\mathbf{X} = \mathbf{R^{-1}Q^\top Y}$ \\
\STATE $\mathbf{s} = \mathbf{X}$ \\
\FOR { $i = 1,2,3,\dots$ }
	\STATE $\mathbf{s} =
        -(\mathbf{M}+\lambda\mathbb{I})^{-1}\mathbf{s}$\\
	\STATE $\mathbf{X}=\mathbf{X} + (-1)^i {\color{red}\mathbf{s}
        \lambda^i}$
\ENDFOR
\end{algorithmic}
\caption{Algorithm for handling rank deficient matrices}
\label{alg:coleman}
\end{algorithm}

The algorithm~\ref{alg:coleman} solves equation~(\ref{eq:taylor}). However, this
version is unstable since typically $\|\mathbf{s}\|$ is very large and
$\lambda^i$ is very small ($\lambda$ is small).

The following algorithm shows a more stable version of
algorithm~\ref{alg:coleman}.

\begin{algorithm}[H]
\begin{algorithmic}[1]
\REQUIRE $\,$ \\
$\mathbf{A}$: design matrix \\
$\mathbf{Y}$: response matrix \\
$\lambda$: rank deficient parameter \\
\ENSURE  $\,$ \\
$\mathbf{X}$: parameters \\
\STATE $\mathbf{M}=\mathbf{A^\top A}$ \\
\STATE Initialize $\mathbf{Q R}=\mathbf{M}$ \\
\STATE $\mathbf{X} = \mathbf{R^{-1}Q^\top Y}$ \\
\STATE $\mathbf{t} = \mathbf{X}$ \\
\FOR { $i = 1,2,3,\dots$ }
        \STATE $\mathbf{t} =\lambda \mathbf{t}$  
        \STATE $\mathbf{t} =  -(\mathbf{M}+\lambda\mathbb{I})^{-1}\mathbf{t}$
	\STATE $\mathbf{X}=\mathbf{X} + \mathbf{t}$
\ENDFOR
\end{algorithmic}
\caption{Algorithm for handling rank deficient matrices improved}
\label{alg:colemanimproved}
\end{algorithm}

Both algorithms are equivalent, but algorithm~\ref{alg:colemanimproved} is more
stable and converges in typically less than 10 steps.
It is important to notice that the QR factorisation is computed only once and it
is computationally less expensive than the SVD.
